---
title: "wrangling-and-cleaning"
author: "Kelsey Ranjbar"
format: html
---

## Load Libraries
```{r}
library(dplyr)
library(tidyr)
library(readr)
```

## Load Data
bit.ly/data-wrangle-lesson-data
```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

## Data Cleaning & Wrangling

### Initial Data Exploration

```{r}
colnames(catch_original)
```

```{r}
head(catch_original)
```

```{r}
summary(catch_original)
```

```{r}
unique(catch_original$Region)
```

```{r}
#View(catch_original)
```

## Pipe Operator
shortcut: Ctrl+Shift+M

GOAL: remove `All` and `notersRedCode` columns

```{r}
catch_data <- catch_original %>% 
  select("Region","Year","Chinook","Sockeye","Coho","Pink","Chum")

catch_data <- catch_original %>% 
  select(-All, -notesRegCode)
```

## Quality Check
```{r}
dplyr::glimpse(catch_data)
```

## Change column content

GOAL: convert the Chinook catch values to `numeric` and overwrite the old Chinood column
```{r}
catch_clean <- catch_data %>% 
  mutate(Chinook = as.numeric(Chinook))
```
## Find the row with the `NA` value

```{r}
which(is.na(catch_clean$Chinook))
  
```

Rewrite this with the pipe operator:
```{r}
i <- catch_clean$Chinook %>% 
  is.na() %>% 
  which()

```

```{r}
catch_data[i,]
```
  
GOAL: change the value of the `Chinook column to 1 if the value is equal to the letter I.

```{r}
catch_clean <- catch_data %>% 
  mutate(Chinook = if_else(condition = Chinook == "I",
                           true = "1",
                           false = Chinook),
         Chinook = as.numeric(Chinook))

```

```{r}
catch_clean[i,]
```
## `pivot_longer()` and `pivot_wider()`

GOAL: transform data from wide format to a long format

```{r}
catch_long <- catch_clean %>% 
  #pivot longer all columns except region and year
  pivot_longer(
    cols = -c(Region, Year),
    names_to = "species",
    values_to = "catch"
  )
```

```{r}
catch_wide <- catch_long %>% 
  pivot_wider(names_from = species,
              values_from = catch)
```

## Renaming columns

GOAL: rename the `catch` column to be called `catch_thousands`
```{r}
catch_long <- catch_long %>% 
  rename(catch_thousands = catch)

head(catch_long)
```

## Adding columns with `mutate()`
GOAL: create a column `catch` with units of fish, instead of thousands of fish
```{r}
catch_long <- catch_long %>% 
  mutate(catch = catch_thousands * 1000)
```

GOAL: remove `catch_thousands` column
```{r}
catch_long <- catch_long %>% 
  select(-catch_thousands)

head(catch_long)
```

## Summary Statistics

GOAL: getting the average catch per region

bit.ly/3ppGUsQ
```{r}
mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch))
  
```

## `filter()` 

GOAL: go back to our `catch_long` data frame and select all the rows from the SSE region
```{r}
SSE_catch <- catch_long %>% 
  filter(Region == "SSE")
```

Quick exercise:
1. filter to catches over one million fish
```{r}
catch_1mil <- catch_long %>% 
  filter(catch > 1000000)
```

2. select rows from the SSE region that have information about Chinook salmon
```{r}
SSE_Chinook <- catch_long %>% 
  filter(Region == "SSE" & species == "Chinook")
```

## Sorting data
GOAL: re-calculate mean catch by region and sort the output by mean catch 
```{r}
mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  arrange(mean_catch)
```

Order in descending order
```{r}
mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))
```

## `separate()` and `unite()`

```{r}
sites_df <- data.frame(site = c("HAW-101",
                                "HAW-103",
                                "OAH-320",
                                "OAH-219",
                                "MAI-039"))
sites_df <- sites_df %>% 
  separate(site, c("island","site_number"), "-")
```

```{r}
sites_df <- sites_df %>% 
  unite(code, island, site_number, sep = "-")
```

## Grand finale
```{r}
catch_original <- read_csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", 
                               method = "libcurl"))

mean_region <- catch_original %>%
  select(-All, -notesRegCode) %>% 
  mutate(Chinook = ifelse(Chinook == "I", 1, Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year), 
               names_to = "species", 
               values_to = "catch") %>%
  mutate(catch = catch*1000) %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))

head(mean_region)
```

